Future Directions

### 2021-02-25 manuscripts/laplacian_smoothing/AISTATS/minimax_optimal_laplacian_smoothing.tex
Added one line
Corrected a few minor typos for AISTATS camera-ready version.

\begin{abstract}
	In this paper we study the statistical properties of Laplacian smoothing, a graph-based approach to nonparametric regression. Under standard regularity conditions, we establish upper bounds on the error of the Laplacian smoothing estimator \smash{$\wh{f}$}, and a goodness-of-fit test also based on \smash{$\wh{f}$}. These upper bounds match the minimax optimal estimation and testing rates of convergence over the first-order Sobolev class $H^1(\Xset)$, for $\Xset \subseteq \Rd$ and $1 \leq d < 4$; in the estimation problem, for $d = 4$, they are optimal modulo a $\log n$ factor. Additionally, we prove that Laplacian smoothing is manifold-adaptive: if $\Xset \subseteq \R^d$ is an $m$-dimensional manifold with $m < d$, then the error rate of Laplacian smoothing (in either estimation or testing) depends only on $m$, in the same way it would if $\Xset$ were a full-dimensional set in $\Reals^m$.
\end{abstract}

@@

For these reasons, a body of work has emerged that analyzes the statistical properties of Laplacian smoothing, and graph-based methods more generally. Roughly speaking, this work can be divided into two categories, based on the perspective they adopt. 
\begin{itemize}
	\item \emph{Fixed design perspective.} Here one treats the design points $X_1,\ldots,X_n$ and the graph $G$ as fixed, and carries out inference on $f_0(X_i)$, $i=1,\ldots,n$. In this problem setting, tight upper bounds have been derived on the error of various graph-based methods (e.g., \citet{wang2016,hutter2016,sadhanala16,sadhanala17,kirichenko2017,kirichenko2018}) and tests (e.g., \citet{sharpnack2010identifying,sharpnack2013b,sharpnack2013,sharpnack2015}), which certify that such procedures are \emph{optimal} over ``function'' classes (in parentheses because these classes really model the $n$-dimensional vector of evaluations). The upside of this work is its generality: in this setting $G$ need not be a geometric graph, but in principle it could be any graph over $V=\{1,\ldots,n\}$. The downside is that, in the context of nonparametric regression, it is as not natural to think of the evaluations of $f_0$ as exhibiting smoothness over some fixed pre-defined graph $G$, and more natural to speak of the smoothness of the function $f_0$ itself. 
	\item \emph{Random design perspective.} Here one treats the design points $X_1,\ldots,X_n$ as independent samples from some distribution $P$ supported on a domain $\Xset \subseteq \Rd$. Inference is drawn on the regression function $f_0: \Xset \to \Reals$, which is typically assumed to be smooth in some \emph{continuum} sense, e.g., it possesses a first derivative bounded in $L^{\infty}$ (H\"{o}lder) or $\Leb^2$ (Sobolev) norm. To conduct graph-based inference, the user first builds a neighborhood graph over the random design points---so that $W_{ij}$ is large when $X_i$ and $X_j$ are close in (say) Euclidean distance---and then computes e.g., \eqref{eqn:laplacian_smoothing} or \eqref{eqn:laplacian_smoothing_test}. In this context, various graph-based procedures have been shown to be \emph{consistent}: as $n \to \infty$, they converge to a continuum limit (see \citet{belkin07,vonluxburg2008,trillos2018} among others). However, until recently such statements were not accompanied by error rates, and even so, such error rates as have been proved \citep{lee2016,trillos2020} are not optimal over continuum function spaces, such as H\"{o}lder or Sobolev classes. 
\end{itemize}
The random design perspective bears a more natural connection with nonparametric regression (the focus in this paper), as it allows us to formulate smoothness based on $f_0$ itself (how it behaves as a continuum function, and not just its evaluations at the design points). In this paper, we will adopt the random design perspective, and seek to answer the following question:
\begin{quote}{When $f_0$ is assumed to smooth in a continuum sense, does Laplacian smoothing achieve optimal performance for estimation and goodness-of-fit testing?} 
\end{quote}
This is no small question---arguably, it is \emph{the} central question of nonparametric regression---and without an answer one cannot fully compare the statistical properties of Laplacian smoothing to alternative methods. It also seems difficult to answer: as we discuss next, there is a fundamental gap between the \emph{discrete} smoothness imposed by the penalty $f^\top \Lap f$ in problem \eqref{eqn:laplacian_smoothing} and the \emph{continuum} smoothness assumed on $f_0$, and in order to obtain sharp upper bounds we will need to bridge this gap in a suitable sense.

@@

By contrast, smoothing splines are optimal only when $d=1$. When $d>1$, thin-plate spline estimator \eqref{eqn:thin_plate_spline} is not even well-posed, in the following sense: for any $(X_1,Y_1),\ldots,(X_n,Y_n)$ and any $\delta>0$, there exists (e.g., \citet{green93} give a construction using ``bump'' functions) a differentiable function $f$ such that $f(X_i) = Y_i$, $i=1,\ldots,n$, and
\begin{equation*}
\int_{\Xset} \|\nabla f(x)\|_2^2 \leq \delta.
\end{equation*}
i.e., $f$ achieves perfect (zero) data loss and arbitrarily small penalty in the problem \eqref{eqn:thin_plate_spline}. This will clearly not lead to a consistent estimator of $f_0$ across the design points (as it always yields $Y_i$ at each $X_i$). In this light, our results for $d>1$ favorably distinguish Laplacian smoothing from its natural variational analog.

@@

Under conditions~\ref{asmp:domain}, \ref{asmp:density}, the minimax estimation rate over a Sobolev ball of radius $M \geq n^{-1/2}$ is (e.g., \citet{tsybakov2008_book}):
\begin{equation}
\label{eqn:sobolev_space_estimation_minimax_rate}
\inf_{\wh{f}} \sup_{f_0 \in H^1(\Xset, M)} \Ebb\Bigl[\norm{\wh{f} - f_0}_{L^2(\Xset)}^2\Bigr] \asymp M^{2d/(2 + d)} n^{-2/(2 + d)}.
\end{equation}
% RJT: This is a bit of a weird reference, because (as per my understanding) Tsybakov only proves the rate is tight for Holder classes, which are contained in the Sobolev class. So technically from Tsybakov alone we only know \gtrsim, not \asymp for this one. It's fine, I wouldn't change anything because it would involve explaining a lot more, I'm just pointing this out.
(Throughout we assume $M \geq n^{-1/2}$---as otherwise the minimax rate is just the parametric rate $n^{-1}$.)

@@

To summarize: for $d = 1,2,$ or $3$, with high probability, the Laplacian smoothing estimator \smash{$\wh{f}$} has in-sample mean squared error that is within a constant factor of the minimax error. Some remarks:
\begin{itemize}
  \item The first-order Sobolev space $H^1(\Xset)$ does not continuously embed into $C^0(\Xset)$ when $d>1$ (in general, the $k$th order space $H^k(\Xset)$ does not continuously embed into $C^0(\Xset)$ except if $2k>d$). For this reason, one really cannot speak of pointwise evaluation of a Sobolev function $f_0 \in H^1(\Xset)$ when $d>1$ (as we do in Theorem~\ref{thm:laplacian_smoothing_estimation1} by defining our target of estimation to be $f_0(X_i)$, $i=1,\ldots,n$). We can resolve this by appealing to what are known as \emph{Lebesgue points}, as explained in the supplement.
% RJT: Alden, please add a section to the supplement to explain this part.
% AJG: Added.
	\item The assumption $M \leq n^{1/d}$ guarantees that the bound in~\eqref{eqn:laplacian_smoothing_estimation1} is meaningful, i.e. that the right hand side of~\eqref{eqn:laplacian_smoothing_estimation1} is of at most a constant order.
	\item The lower bound on $r$ imposed in condition \ref{asmp:ls_kernel_radius_estimation} is compatible with practice, where by far the most common choice of radius is the connectivity threshold \smash{$r \asymp (\log(n)/n)^{1/d}$}, which makes \smash{$G_{n,r}$} as sparse as possible while still being connected, for maximum computational efficiency. The upper bound may seem a bit more mysterious---we need it for technical reasons to ensure that \smash{$\wh{f}$} does not overfit, but note that as a practical matter one rarely chooses $r$ to be so large anyway.
	\item It is possible to extend \smash{$\wh{f}$} to be defined on all of $\Xset$ and then evaluate the error of such an extension (as measured against $f_0$) in $\Leb^2(\Xset)$ norm. When \smash{$\wh{f}$} and $f_0$ are suitably smooth, tools from empirical process theory (see e.g., Chapter 14 of \citet{wainwright2019}) or approximation theory (e.g., Section 15.5 of \citet{johnstone2011}) guarantee that the $\Leb^2(\Xset)$ error is not too much greater than its in-sample counterpart. In fact, as shown in the supplement, if $f_0$ is Lipschitz smooth and we extend \smash{$\wh{f}$} to be piecewise constant over the Voronoi tesselation induced by $X_1,\ldots,X_n$, then the out-of-sample error \smash{$\|\wh{f}-f_0\|_{\Leb^2(\Xset)}$} is within a negligible factor of the in-sample error \smash{$\|\wh{f}-f_0\|_n$}. We leave analysis of the Sobolev case to future work.
	\item Additionally, under the stronger assumption that $f_0$ is Lipschitz smooth, we can replace the factor of $\delta$ in the high probability bound by the sharper $\delta^2/n$.
\end{itemize}

When $d = 4$, our analysis results in an upper bound for the error of Laplacian smoothing that is within a \smash{$(\log n)^{1/3}$} factor of the minimax error rate. But when $d \geq 5$, our upper bounds are much worse.
\begin{theorem}
	\label{thm:laplacian_smoothing_estimation2}
  Under the assumptions of Theorem~\ref{thm:laplacian_smoothing_estimation1}, if $\Xset$ has dimension $d = 4$, \smash{$r \asymp (\log n/n)^{1/4}$} and \smash{$\rho = M^{-2/3}(nr^6)^{-1}(\log n/n)^{1/3}$}, then we obtain 
	\begin{equation*}
	\bigl\|\wh{f} - f_0\bigr\|_n^2 \leq \frac{C}{\delta} M^{4/3} \biggl(\frac{\log n}{n}\biggr)^{1/3},
	\end{equation*}
with the same probability guarantee as in Theorem~\ref{thm:laplacian_smoothing_estimation1}. If the dimension of $\Xset$ is $d \geq 5$, \smash{$r \asymp (\log n/n)^{1/d}$} and \smash{$\rho =  M^{-2/3}(nr^{2 + d})^{-1}n^{-4/(3d)}$}, then
	\begin{equation*}
	\bigl\|\wh{f} - f_0\bigr\|_n^2 \leq \frac{C}{\delta} M^{4/3} \biggl(\frac{\log n}{n}\biggr)^{-4/(3d)},
	\end{equation*}
% RJT: Alden, please verify what I wrote is correct, and fill in the result for d \geq 5.
% AJG: Verified and filled in.
\end{theorem}
This mirrors the conclusions of \citet{sadhanala16} who investigate estimation rates of Laplacian smoothing over the $d$-dimensional grid graph. These authors argue that their analysis is tight, and that it is likely the estimator, not the analysis, that is deficient when $d \geq 5$. Formalizing such a claim turns out to be harder in the random design setting than in the fixed design setting, and we leave it for future work.

@@

Of course, we know this is not the case: $\wh{f}$ and $\wt{f}$ look very differently when $d > 1$. What is driving this difference? The key point is that the discretization imposed by the graph $G_{n,r}$---which might seem problematic at first glance---turns out to be a blessing. The problem with~\eqref{eqn:thin_plate_spline} is that the class $H^1(\Xset)$, which fundamentally underlies the criterion, is far ``too big'' for $d > 1$. This is meant in various related senses. By the Sobolev embedding theorem, for $d>1$, the class $H^1(\Xset)$ does not continuously embed into any H\"{o}lder space; and in fact it does not even continuously embed into $C^0(\Xset)$. Thus we cannot really restrict the optimization to \emph{continuous} and weakly differentiable functions, as we could when $d=1$ (the smoothing spline case), without throwing out a substantial subset of functions in $H^1(\Xset)$. Even among continuous and differentiable functions $f$, as we explained previously, we can use ``bump'' functions (as in \citet{green93}) to construct $f$ that interpolates the pairs $(X_i,Y_i)$, $i=1,\ldots,n$ and achieves arbitrarily small penalty (and hence criterion) in \eqref{eqn:thin_plate_spline}. In this sense, any estimator resulting from solving \eqref{eqn:thin_plate_spline} estimator will clearly be inconsistent.

>>>>

\begin{abstract}
	In this paper we study the statistical properties of Laplacian smoothing, a graph-based approach to nonparametric regression. Under standard regularity conditions, we establish upper bounds on the error of the Laplacian smoothing estimator \smash{$\wh{f}$}, and a goodness-of-fit test also based on \smash{$\wh{f}$}. These upper bounds match the minimax optimal estimation and testing rates of convergence over the first-order Sobolev class $H^1(\Xset)$, for $\Xset \subseteq \Rd$ and $1 \leq d < 4$; in the estimation problem, for $d = 4$, they are optimal modulo a $\log n$ factor. Additionally, we prove that Laplacian smoothing is manifold-adaptive: if $\Xset \subseteq \Reals^d$ is an $m$-dimensional manifold with $m < d$, then the error rate of Laplacian smoothing (in either estimation or testing) depends only on $m$, in the same way it would if $\Xset$ were a full-dimensional set in $\Reals^m$.
\end{abstract}

@@

For these reasons, a body of work has emerged that analyzes the statistical properties of Laplacian smoothing, and graph-based methods more generally. Roughly speaking, this work can be divided into two categories, based on the perspective they adopt. 
\begin{itemize}
	\item \emph{Fixed design perspective.} Here one treats the design points $X_1,\ldots,X_n$ and the graph $G$ as fixed, and carries out inference on $f_0(X_i)$, $i=1,\ldots,n$. In this problem setting, tight upper bounds have been derived on the error of various graph-based methods (e.g., \citet{wang2016,hutter2016,sadhanala16,sadhanala17,kirichenko2017,kirichenko2018}) and tests (e.g., \citet{sharpnack2010identifying,sharpnack2013b,sharpnack2013,sharpnack2015}), which certify that such procedures are \emph{optimal} over ``function'' classes (in quotes because these classes really model the $n$-dimensional vector of evaluations). The upside of this work is its generality: in this setting $G$ need not be a geometric graph, but in principle it could be any graph over $V=\{1,\ldots,n\}$. The downside is that, in the context of nonparametric regression, it is not as natural to think of the evaluations of $f_0$ as exhibiting smoothness over some fixed pre-defined graph $G$, and more natural to speak of the smoothness of the function $f_0$ itself. 
	\item \emph{Random design perspective.} Here one treats the design points $X_1,\ldots,X_n$ as independent samples from some distribution $P$ supported on a domain $\Xset \subseteq \Rd$. Inference is drawn on the regression function $f_0: \Xset \to \Reals$, which is typically assumed to be smooth in some \emph{continuum} sense, e.g., it possesses a first derivative bounded in $L^{\infty}$ (H\"{o}lder) or $\Leb^2$ (Sobolev) norm. To conduct graph-based inference, the user first builds a neighborhood graph over the random design points---so that $W_{ij}$ is large when $X_i$ and $X_j$ are close in (say) Euclidean distance---and then computes e.g., \eqref{eqn:laplacian_smoothing} or \eqref{eqn:laplacian_smoothing_test}. In this context, various graph-based procedures have been shown to be \emph{consistent}: as $n \to \infty$, they converge to a continuum limit (see \citet{belkin07,vonluxburg2008,trillos2018} among others). However, until recently such statements were not accompanied by error rates, and even so, such error rates as have been proved \citep{lee2016,trillos2020} are not optimal over continuum function spaces, such as H\"{o}lder or Sobolev classes. 
\end{itemize}
The random design perspective bears a more natural connection with nonparametric regression (the focus in this paper), as it allows us to formulate smoothness based on $f_0$ itself (how it behaves as a continuum function, and not just its evaluations at the design points). In this paper, we will adopt the random design perspective, and seek to answer the following question:
\begin{quote}{When $f_0$ is assumed to be smooth in a continuum sense, does Laplacian smoothing achieve optimal performance for estimation and goodness-of-fit testing?} 
\end{quote}
This is no small question---arguably, it is \emph{the} central question of nonparametric regression---and without an answer one cannot fully compare the statistical properties of Laplacian smoothing to alternative methods. It also seems difficult to answer: as we discuss next, there is a fundamental gap between the \emph{discrete} smoothness imposed by the penalty $f^\top \Lap f$ in problem \eqref{eqn:laplacian_smoothing} and the \emph{continuum} smoothness assumed on $f_0$, and in order to obtain sharp upper bounds we will need to bridge this gap in a suitable sense.

@@

By contrast, smoothing splines are optimal only when $d=1$. When $d>1$, thin-plate spline estimator \eqref{eqn:thin_plate_spline} is not even well-posed, in the following sense: for any $(X_1,Y_1),\ldots,(X_n,Y_n)$ and any $\delta>0$, there exists (e.g., \citet{green93} give a construction using ``bump'' functions) a differentiable function $f$ such that $f(X_i) = Y_i$, $i=1,\ldots,n$, and
\begin{equation*}
\int_{\Xset} \|\nabla f(x)\|_2^2 \leq \delta.
\end{equation*}
In other words, $f$ achieves perfect (zero) data loss and arbitrarily small penalty in the problem \eqref{eqn:thin_plate_spline}. This will clearly not lead to a consistent estimator of $f_0$ across the design points (as it always yields $Y_i$ at each $X_i$). In this light, our results for $d>1$ favorably distinguish Laplacian smoothing from its natural variational analog.

@@

Under conditions~\ref{asmp:domain}, \ref{asmp:density}, the minimax estimation rate over a Sobolev ball of radius $M \geq n^{-1/2}$ is (e.g., \citet{tsybakov2008_book}):
\begin{equation}
\label{eqn:sobolev_space_estimation_minimax_rate}
\inf_{\wh{f}} \sup_{f_0 \in H^1(\Xset, M)} \Ebb\Bigl[\norm{\wh{f} - f_0}_{L^2(\Xset)}^2\Bigr] \asymp M^{2d/(2 + d)} n^{-2/(2 + d)}.
\end{equation}
% RJT: This is a bit of a weird reference, because (as per my understanding) Tsybakov only proves the rate is tight for Holder classes, which are contained in the Sobolev class. So technically from Tsybakov alone we only know \gtrsim, not \asymp for this one. It's fine, I wouldn't change anything because it would involve explaining a lot more, I'm just pointing this out.
(Throughout we assume $M \geq n^{-1/2}$, as otherwise the minimax rate is just the parametric rate $n^{-1}$.)

@@

To summarize: for $d = 1,2,$ or $3$, with high probability, the Laplacian smoothing estimator \smash{$\wh{f}$} has in-sample mean squared error that is within a constant factor of the minimax error. Some remarks:
\begin{itemize}
  \item The first-order Sobolev space $H^1(\Xset)$ does not continuously embed into $C^0(\Xset)$ when $d>1$ (in general, the $k$th order space $H^k(\Xset)$ does not continuously embed into $C^0(\Xset)$ except if $2k>d$). For this reason, one really cannot speak of pointwise evaluation of a Sobolev function $f_0 \in H^1(\Xset)$ when $d>1$ (as we do in Theorem~\ref{thm:laplacian_smoothing_estimation1} by defining our target of estimation to be $f_0(X_i)$, $i=1,\ldots,n$). We can resolve this by appealing to what are known as \emph{Lebesgue points}, as explained in the supplement.
% RJT: Alden, please add a section to the supplement to explain this part.
% AJG: Added.
	\item The assumption $M \leq n^{1/d}$ guarantees that the bound in~\eqref{eqn:laplacian_smoothing_estimation1} is meaningful, i.e. that the right hand side of~\eqref{eqn:laplacian_smoothing_estimation1} is of at most a constant order.
	\item The lower bound on $r$ imposed in condition \ref{asmp:ls_kernel_radius_estimation} is compatible with practice, where by far the most common choice of radius is the connectivity threshold \smash{$r \asymp (\log(n)/n)^{1/d}$}, which makes \smash{$G_{n,r}$} as sparse as possible while still being connected, for maximum computational efficiency. The upper bound may seem a bit more mysterious---we need it for technical reasons to ensure that \smash{$\wh{f}$} does not overfit, but note that as a practical matter one rarely chooses $r$ to be so large anyway.
	\item It is possible to extend \smash{$\wh{f}$} to be defined on all of $\Xset$ and then evaluate the error of such an extension (as measured against $f_0$) in $\Leb^2(\Xset)$ norm. When \smash{$\wh{f}$} and $f_0$ are suitably smooth, tools from empirical process theory (see e.g., Chapter 14 of \citet{wainwright2019}) or approximation theory (e.g., Section 15.5 of \citet{johnstone2011}) guarantee that the $\Leb^2(\Xset)$ error is not too much greater than its in-sample counterpart. In fact, as shown in the supplement, if $f_0$ is Lipschitz smooth and we extend \smash{$\wh{f}$} to be piecewise constant over the Voronoi tesselation induced by $X_1,\ldots,X_n$, then the out-of-sample error \smash{$\|\wh{f}-f_0\|_{\Leb^2(\Xset)}$} is within a negligible factor of the in-sample error \smash{$\|\wh{f}-f_0\|_n$}. We leave analysis of the Sobolev case to future work.
	\item Under the assumption that $f_0$ is Lipschitz smooth, we can also replace the factor of $\delta$ in the high probability bound by the sharper $\delta^2/n$.
\end{itemize}

When $d = 4$, our analysis results in an upper bound for the error of Laplacian smoothing that is within a \smash{$(\log n)^{1/3}$} factor of the minimax error rate. But when $d \geq 5$, our upper bounds are much worse.
\begin{theorem}
	\label{thm:laplacian_smoothing_estimation2}
  Under the assumptions of Theorem~\ref{thm:laplacian_smoothing_estimation1}, if $\Xset$ has dimension $d = 4$, \smash{$r \asymp (\log n/n)^{1/4}$} and \smash{$\rho = M^{-2/3}(nr^6)^{-1}(\log n/n)^{1/3}$}, then we obtain 
	\begin{equation*}
	\bigl\|\wh{f} - f_0\bigr\|_n^2 \leq \frac{C}{\delta} M^{4/3} \biggl(\frac{\log n}{n}\biggr)^{1/3},
	\end{equation*}
with the same probability guarantee as in Theorem~\ref{thm:laplacian_smoothing_estimation1}. If the dimension of $\Xset$ is $d \geq 5$, \smash{$r \asymp (\log n/n)^{1/d}$} and \smash{$\rho =  M^{-2/3}(nr^{2 + d})^{-1}n^{-4/(3d)}$}, then
	\begin{equation*}
	\bigl\|\wh{f} - f_0\bigr\|_n^2 \leq \frac{C}{\delta} M^{4/3} \biggl(\frac{\log n}{n}\biggr)^{-4/(3d)},
	\end{equation*}
again with the same probability guarantee.
% RJT: Alden, please verify what I wrote is correct, and fill in the result for d \geq 5.
% AJG: Verified and filled in.
\end{theorem}
This mirrors the conclusions of \citet{sadhanala16} who investigate estimation rates of Laplacian smoothing over the $d$-dimensional grid graph. These authors argue that their analysis is tight, and that it is likely the estimator, not the analysis, that is deficient when $d \geq 5$. Formalizing such a claim turns out to be harder in the random design setting than in the fixed design setting, and we leave it for future work.

@@

Of course, we know this is not the case: $\wh{f}$ and $\wt{f}$ look very different when $d > 1$. What is driving this difference? The key point is that the discretization imposed by the graph $G_{n,r}$---which might seem problematic at first glance---turns out to be a blessing. The problem with~\eqref{eqn:thin_plate_spline} is that the class $H^1(\Xset)$, which fundamentally underlies the criterion, is far ``too big'' for $d > 1$. This is meant in various related senses. By the Sobolev embedding theorem, for $d>1$, the class $H^1(\Xset)$ does not continuously embed into any H\"{o}lder space; and in fact it does not even continuously embed into $C^0(\Xset)$. Thus we cannot really restrict the optimization to \emph{continuous} and weakly differentiable functions, as we could when $d=1$ (the smoothing spline case), without throwing out a substantial subset of functions in $H^1(\Xset)$. Even among continuous and differentiable functions $f$, as we explained previously, we can use ``bump'' functions (as in \citet{green93}) to construct $f$ that interpolates the pairs $(X_i,Y_i)$, $i=1,\ldots,n$ and achieves arbitrarily small penalty (and hence criterion) in \eqref{eqn:thin_plate_spline}. In this sense, any estimator resulting from solving \eqref{eqn:thin_plate_spline} will clearly be inconsistent.
